---
title: "Forecasting"
author: "Gabriel Griffin"
date: "2026-09-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(lubridate)
library(fpp3)

kit_data <- read_csv("kit_production.csv")

names(kit_data)
head(kit_data)
summary(kit_data)
head(kit_data$Date, 10)
```

```{r}
agg_demand <- kit_data %>%
  mutate(
    Date = ymd(Date),
    yw = yearweek(Date)
  ) %>%
  group_by(yw) %>%
  summarise(
    demand = sum(Demand),
    .groups = "drop"
  ) %>%
  as_tsibble(index = yw)
```

```{r}
agg_demand%>%
  autoplot()
names(agg_demand)
```
Extreme seasonality and overall trend down
```{r}
monthly_demand <- kit_data %>%
  mutate(
    Date = ymd(Date),
    ym = yearmonth(Date)
  ) %>%
  group_by(ym) %>%
  summarise(
    demand = sum(Demand),
    .groups = "drop"
  ) %>%
  as_tsibble(index = ym)

monthly_demand %>%
  autoplot(demand)
```
```{r}
train_data <- agg_demand %>%
  filter(year(yw) < 2023)

test_data <- agg_demand %>%
  filter(year(yw) >= 2023)

library(randtests)

runs.test(train_data$demand)

acf(train_data$demand)
```
Autocorrelation with span 7, p values < 0.05, demand isnt random.

```{r}
dat_mods <- train_data %>%
  model(
    MA = ARIMA(
      demand ~ 0 + pdq(
        p = 0,
        d = 0,
        q = 2
      )
    ),

    ARn = ARIMA(
      demand ~ pdq(
        p = 4,
        d = 0,
        q = 0
      )
    )
  )
```

```{r}
dat_mods %>%
  report() %>%
  arrange(AIC)
```

Lower AIC for Arn, but not by much, however AICc and BIC favor MA. Final model selection should be based on out of sample forecast accuracy. AR(4)

```{r}
dat_mods %>%
  forecast(new_data = test_data) %>%
  accuracy(test_data) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)
```
Error lower in all metrics for ARn.

```{r}
dat_mods %>%
  forecast(new_data = test_data) %>%
  autoplot(test_data,
           level = 80,
           size = 0.5) +
  geom_line(
    aes(y = demand),
    data = agg_demand
  )
```
Look similar with ARn appearing as a lower bound for MA.

```{r}
dat_mods_rf <- dat_mods %>%
  refit(new_data = agg_demand)

dat_mods_rf %>%
  accuracy() %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)
```
Refitted shows ARn better in all error metrics with MAE significantly better.

```{r}
agg_demand %>%
  filter(year(yw) >= 2023) %>%
  autoplot(demand) +
  geom_line(
    aes(
      y = .fitted,
      colour = .model
    ),
    data = dat_mods_rf %>%
      fitted()
  )
```
Want to use holt winters because of reasons stated earlier so use monthly because 52>24.

```{r}
monthly_train <- monthly_demand %>%
  filter(year(ym) < 2023)

monthly_test <- monthly_demand %>%
  filter(year(ym) >= 2023)
```

```{r}
acf(monthly_train$demand)
```
Autocorrelated with lag 2

```{r}
monthly_mods <- monthly_train %>%
  model(
    ES = ETS(
      demand ~
        error("A") +
        trend("A") +
        season("A", period = 12)
    )
  )
```

```{r}
monthly_mods %>%
  accuracy() %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)

monthly_mods %>%
  forecast(new_data = monthly_test) %>%
  accuracy(monthly_test) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)

monthly_mods %>%
  forecast(new_data = monthly_test) %>%
  autoplot(monthly_test,
           level = 80,
           size = 0.5) +
  geom_line(
    aes(y = demand),
    data = monthly_demand
  )
```

Cant Compare to weekly directly, but while weekly models ae more useful for production scheduling and inventory, the monthly model is useful for long term trend and seasonality.

```{r}
monthly_mods_rf <- monthly_mods %>%
  refit(new_data = monthly_demand)

monthly_mods_rf %>%
  accuracy() %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)

monthly_demand %>%
  autoplot(demand) +
  geom_line(
    aes(
      y = .fitted,
      colour = .model
    ),
    data = monthly_mods_rf %>%
      fitted()
  )

```
model predicts monthly data very well with error both above and below true.

Actual forecast:

```{r}
final_model <- agg_demand %>%
  model(
    AR4 = ARIMA(
      demand ~ pdq(
        p = 4,
        d = 0,
        q = 0
      )
    )
  )

final_forecast <- final_model %>%
  forecast(h = 52)
```

```{r}

final_forecast %>%
  autoplot(
    agg_demand,
    level = 80
  ) +
  labs(
    x = "Week",
    y = "Demand",
    title = "Weekly Demand Forecast Using AR(4)"
  )
```

Ar 4 was best on the holdout set but cannot be stably estimated on complete data, need stable model.

```{r}
ar_mods <- train_data %>%
  model(
    AR1 = ARIMA(demand ~ pdq(1, 0, 0)),
    AR2 = ARIMA(demand ~ pdq(2, 0, 0)),
    AR3 = ARIMA(demand ~ pdq(3, 0, 0)),
    AR4 = ARIMA(demand ~ pdq(4, 0, 0)),
    AR5 = ARIMA(demand ~ pdq(5, 0, 0)),
    AR6 = ARIMA(demand ~ pdq(6, 0, 0)),
    AR7 = ARIMA(demand ~ pdq(7, 0, 0))
  )

ar_mods %>%
  forecast(new_data = test_data) %>%
  accuracy(test_data) %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)
```

AR3 better than AR4 actually.

```{r}
final_model <- agg_demand %>%
  model(
    AR3 = ARIMA(
      demand ~ pdq(
        p = 3,
        d = 0,
        q = 0
      )
    )
  )
```

Still getting numerical instability errors over testing data.

```{r}

full_mods <- agg_demand %>%
  model(
    AR1 = ARIMA(demand ~ pdq(1, 0, 0)),
    AR2 = ARIMA(demand ~ pdq(2, 0, 0)),
    AR3 = ARIMA(demand ~ pdq(3, 0, 0)),
    AR4 = ARIMA(demand ~ pdq(4, 0, 0)),
    AR5 = ARIMA(demand ~ pdq(5, 0, 0)),
    AR6 = ARIMA(demand ~ pdq(6, 0, 0)),
    
    MA1 = ARIMA(demand ~ 0 + pdq(0, 0, 1)),
    MA2 = ARIMA(demand ~ 0 + pdq(0, 0, 2)),
    MA3 = ARIMA(demand ~ 0 + pdq(0, 0, 3))
  )

full_mods %>%
  report()

```

AR2 - AR4 have instability issues, use AR 5 as almost as good for error but stable.

```{r}
final_model <- agg_demand %>%
  model(
    AR5 = ARIMA(
      demand ~ pdq(
        p = 5,
        d = 0,
        q = 0
      )
    )
  )

final_model %>%
  report()
```

```{r}
final_forecast <- final_model %>%
  forecast(h = 52)

final_forecast %>%
  autoplot(
    agg_demand,
    level = 80
  ) +
  labs(
    x = "Week",
    y = "Demand",
    title = "Weekly Demand Forecast Using AR(5)"
  )
```
Doesn't seem to follow overall trend, but matches seasonality very well. I would use holt winters monthly to gauge overall trend then using only local behavior of AR5 adjust holt winters forecast.

```{r}
monthly_mods_rf <- monthly_mods %>%
  refit(new_data = monthly_demand)

hw_forecast <- monthly_mods_rf %>%
  forecast(h = 12)

hw_forecast %>%
  autoplot(
    monthly_demand,
    level = 80
  ) +
  labs(
    x = "Month",
    y = "Demand",
    title = "Monthly Demand Forecast Using Holt-Winters"
  )
```


